Open Policing Project Tutorial

### 2019-03-11 tutorial.Rmd
progress

@@

## Load the data
# Replace the path below with the path to where your data lives
data_path <- "~/opp/data/states/pa/philadelphia/clean/philadelphia.rds"
stops <- read_rds(data_path)$data
# data_path <- "~/Desktop/pa_philadelphia_2019_02_25.rds"
# stops <- read_rds(data_path)

@@

```{r}
stops <- stops %>% filter(year(date) < 2018)
```

@@

**tidyverse tip:** The "pipe" operator, `%>%`, helps to keep our code clean. It
just places the previous item into the first argument of the function. So,
`x %>% f()` is simply `f(x)`. While in a one-function call the pipe might feel
silly and unnecessary, it's going to become _really_ helpful once we start
wanting to do multiple transformations to our data. For instance, that means
that instead of typing `nrows(filter(stops, year < 2018))` we can write
`stops %>% filter(year < 2018) %>% nrows()`, which is easier to understand.

@@

We used ``ggplot`` to make this plot. This package provides a lot of options for
customization, and there are many tweaks we could make to this plot to make it
nicer. It's syntax is a little complicated, though, and in our data
exploration, it's enough just to get a coarse visual like this.

@@

**Fun fact:** This same trend does _not_ hold true for pedestrian stops. In fact,
if we hadn't filtered to only vehicular stops, we wouldn't have noticed this,
because this same plot over the combined pedestrian + vehicular data looks
unremarkable. The takeaway is that looking at trends by sub-categories can often
be very helpful. (e.g., In Nashville, looking at the different listed stop 
reasons uncovers the extent of the disparities.)

@@

One thing that we often want to do with data is disaggregate it. That is, we
might want to take the data and break it down into smaller subpopulations. Then,
when we ask questions, we can ask about each piece---for instance, each
demographic group, or each police district---instead asking about the population
as a whole.

The way to do this in `R` is with `group_by()` and summarise. The standard way
to use `group_by()` is to call `group_by(DATA, COL_NAME)`, where `DATA` is our
dataframe and `COL_NAME` is the name of a column. What `group_by()` then does is
take all the rows in the dataframe `DATA` and put them into different groups,
one for each different value in the column `COL_NAME`. So, for instance, if we
called `group_by(stops, district)`, `R` would hand back to us the `stops`
dataframe with all of its columns broken into different groups, one for each
police district.

@@

Let's put it all together with an example. A straightforward one is: `stops %>%
group_by(subject_race) %>% summarize(arrest_rate = mean(arrest_made, na.rm =
True))`. Typing this into the `R` console will calculates arrest rates
disaggregated by race.

@@

As with the stop rates, we can make a quantitative claim about disparities in
search and frisk rates by dividing the minority rate by the white rate.
Here we see that among drivers who were stopped, blacks were searched at a rate
1.5 times higher than whites were, and Hispanics were searched at a rate 1.2
times higher than whites were. Black drivers were frisked at a rate 2.1 times
higher than white drivers were, and Hispanic drivers were frisked at a rate
1.5 times higher than white drivers were.

@@

While these baseline stats give us a sense that there are racial disparities in
policing practices is Philadelphia, they are not evidence discrimination. The
argument against the banchmark test is that we haven't identified the correct
baseline to compare to.

@@

Search and frisk rates are slightly less suspect, since among the stopped 
population, it's more reasonable to believe that people of different races
offend at equal rates. In the context of searches, this means assuming that
all races exhibit probable cause of possessing contraband at equal rates. And
in the case of frisks, this means assuming that all race exhibit reasonable
articulable suspicion of possessing a weapon at equal rates. Of course, these
assumptions can also be argued against. One could claim that the stopped population
isn't a good measure of the true racial distribution of probable cause.

@@

```{r}
hit_rates <- 
  stops %>% 
  filter(search_conducted) %>% 
  group_by(subject_race, district) %>% 
  summarize(hit_rate = mean(contraband_found, na.rm = T))
```

@@

```{r}
# Reshape table to show hit rates of minorities vs whites
hit_rates <-
  hit_rates %>% 
  filter(subject_race %in% c("black", "white", "hispanic")) %>% 
  spread(subject_race, hit_rate, fill = 0) %>% 
  rename(white_hit_rate = white) %>% 
  gather(minority_race, minority_hit_rate, c(black, hispanic))

@@

**Quick Tip**: You might have noticed a strange symbol in the code above:
`%in%`. This is really just shorthand for the sentence, "the `subject_race` is
`"black"` or its `"white"` or its `"hispanic"`." Without the `%in%` syntax, we
would have to write this in `R` as: `subject_race == "black" | subject_race ==
"white" | subject_race == "hispanic"`. The `%in%` syntax is a lot shorter and
clearer, we think you'll agree.

@@

hit_rates %>% 
  ggplot(aes(
    x = white_hit_rate,
    y = minority_hit_rate
  )) +
  geom_point() +
  # This sets a diagonal reference line (line of equal hit rates)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  # These next few lines just make the axes pretty and even
  scale_x_continuous("White hit rate", 
    limits = c(0, max_hit_rate + 0.01),
    labels = scales::percent
  ) +
  scale_y_continuous("Minority hit rate", 
    limits = c(0, max_hit_rate + 0.01),
    labels = scales::percent
  ) +
  # This makes sure that 1% on the x-axis is the same as 1% on the y-axis
  coord_fixed() +
  # This allows us to compare black v. white and Hispanic v. white side by
  # side, in panels
  facet_grid(cols = vars(minority_race))
```

This is a good start. However, it's hard to tell where the mass is. That is, 
maybe the points above the dotted line (i.e., the points where minority hit
rates are _higher_ than white hit rates), maybe those districts have all of 
Philadelphia's popualtion, and the districts below the line only represent 
a few people. While this is unlikely, it's still good to have a marker of how 
much weight or emphasis to give each of the points on our plot.

@@

hit_rates %>% 
  left_join(
    search_counts, 
    by = c("district", "minority_race")
  ) %>% 
  ggplot(aes(
    x = white_hit_rate,
    y = minority_hit_rate
  )) +
  geom_point(aes(size = num_searches), pch = 21) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_x_continuous("White hit rate", 
    limits = c(0, max_hit_rate + 0.01),
    labels = scales::percent
  ) +
  scale_y_continuous("Minority hit rate", 
    limits = c(0, max_hit_rate + 0.01),
    labels = scales::percent
  ) +
  coord_fixed() +
  facet_grid(cols = vars(minority_race))
```

@@

(Note that we distinguish here between _sunset_ and _dusk_. Sunset is the point 
in time where the sun dips below the horizon. Dusk, also known as the end of 
civil twilight, is the time when the sun is six degrees below the horizon, and 
is it is widely considered to be dark. We'll explain why this is relevant in a
moment.)

@@

One quick note: We want to make sure to filtering out stops that occurred in the
approximately 30-minute period between sunset and dusk (end of civil twilight),
since it is ambiguous as to whether that period is light or dark, which would
muddle up our analysis.

@@

TODO(amyshoe)

>>>>

First, let's load the necessary libraries and data that will allow us to
begin our investigation!

@@

## Load the data
# Replace the path below with the path to where your data lives
data_path <- "~/Desktop/pa_philadelphia_2019_02_25.rds"
stops <- read_rds(data_path)

@@

```{r}
stops <- filter(stops, year(date) < 2018)
```

@@

**tidyverse tip**: The "pipe" operator, `%>%`, helps to keep our code clean. It
just places the previous item into the first argument of the function. So,
`x %>% f()` is simply `f(x)`. While in a one-function call the pipe might feel
silly and unnecessary, it's going to become _really_ helpful once we start
wanting to do multiple transformations to our data. For instance, that means
that instead of typing `nrows(filter(stops, year < 2018))` we can write
`stops %>% filter(year < 2018) %>% nrows()`, which is easier to understand.

@@

We used ``ggplot`` to make this plot. This package provides a lot of options for
customization, and there are many tweaks we could make to this plot to make it
nicer. Its syntax is a little complicated, though, and in our data
exploration, it's enough just to get a coarse visual like this.

@@

**Fun fact:** This same trend does _not_ hold true for pedestrian stops. In fact,
if we hadn't filtered to only vehicular stops, we wouldn't have noticed this,
because this same plot over the combined pedestrian + vehicular data looks
unremarkable. The takeaway is that looking at trends by sub-categories can often
be very helpful. (E.g., in Nashville, looking at the different listed stop 
reasons uncovers the extent of the disparities.)

@@

One thing that we often want to do with data is disaggregate it. That is, we
might want to take the data and break it down into smaller subpopulations. Then,
when we ask questions, we can ask about each piece---for instance, each
demographic group, or each police district---instead of asking about the population
as a whole.

The way to do this in `R` is with `group_by()` and `summarize()`. The standard way
to use `group_by()` is to call `group_by(DATA, COL_NAME)`, where `DATA` is our
dataframe and `COL_NAME` is the name of a column. What `group_by()` then does is
take all the rows in the dataframe `DATA` and put them into different groups,
one for each different value in the column `COL_NAME`. So, for instance, if we
called `group_by(stops, district)`, `R` would hand back to us the `stops`
dataframe with all of its columns broken into different groups, one for each
police district.

@@

Let's put it all together with an example. A straightforward one is: `stops %>%
group_by(subject_race) %>% summarize(arrest_rate = mean(arrest_made))`. Typing this into the `R` console will calculates arrest rates
disaggregated by race.

@@

Note that with functions like `mean()`, if any of the values that are being
averaged are `NA`, the output value of the mean will simply be `NA`. We don't have
to worry about that with our Philly data, because it's super clean. But you may
encounter this problem in some of your own data. The easy workaround is to pass
the argument `na.rm = T` into `mean()` -- as you see in the code above. This tells 
`R` to remove the `NA` values and compute the mean over all the non-NA values.

Now let's dive into these results! As with the stop rates, we can make a 
quantitative claim about disparities in
search and frisk rates by dividing the minority rate by the white rate.
Here we see that among drivers who were stopped, blacks were searched at a rate
1.5 times higher than whites were, and Hispanics were searched at a rate 1.2
times higher than whites were. Black drivers were frisked at a rate 2.1 times
higher than white drivers were, and Hispanic drivers were frisked at a rate
1.5 times higher than white drivers were.

@@

While these baseline stats give us a sense that there are racial disparities in
policing practices in Philadelphia, they are not evidence of discrimination. The
argument against the benchmark test is that we haven't identified the correct
baseline to compare to.

@@

An even simpler critique of the population bechmark for stop rates is that it
doesn't account for possible race-specific differences in driving behavior, including
amount of time spent on the road (and adherence to traffic laws, as mentioned above).
If black drivers, hypothetically, spend more time on the road than whites, that in and
of itself could explain the higher stop rates we see for black drivers, even in the
absence of discrimination.

Search and frisk rates are slightly less suspect, since among the stopped 
population, it's more reasonable to believe that people of different races
offend at equal rates. In the context of searches, this means assuming that
all races exhibit probable cause of possessing contraband at equal rates. And
in the case of frisks, this means assuming that all races exhibit reasonable
articulable suspicion of possessing a weapon at equal rates. Of course, these
assumptions can also be argued against. One could claim that the stopped population
isn't a good measure of the true racial distribution of probable cause.

@@

```{r}
hit_rates <- 
  stops %>% 
  filter(search_conducted) %>% 
  group_by(subject_race, district) %>% 
  summarize(hit_rate = mean(contraband_found, na.rm = T))

@@

```{r}
# Reshape table to show hit rates of minorities vs whites
hit_rates <-
  hit_rates %>% 
  filter(subject_race %in% c("black", "white", "hispanic")) %>% 
  spread(subject_race, hit_rate, fill = 0) %>% 
  rename(white_hit_rate = white) %>% 
  gather(minority_race, minority_hit_rate, c(black, hispanic)) %>%
  arrange(district)

@@

**Quick Tip**: You might have noticed a strange symbol in the code above:
`%in%`. This is really just shorthand for the sentence, "the `subject_race` is
`"black"` or it is `"white"` or it is `"hispanic"`." Without the `%in%` syntax, we
would have to write this in `R` as: `subject_race == "black" | subject_race ==
"white" | subject_race == "hispanic"`. The `%in%` syntax is a lot shorter and
clearer; we think you'll agree.

@@

hit_rates %>% 
  ggplot(aes(
    x = white_hit_rate,
    y = minority_hit_rate
  )) +
  geom_point() +
  # This sets a diagonal reference line (line of equal hit rates)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  # These next few lines just make the axes pretty and even
  scale_x_continuous("White hit rate", 
    limits = c(0, max_hit_rate + 0.01),
    labels = scales::percent
  ) +
  scale_y_continuous("Minority hit rate", 
    limits = c(0, max_hit_rate + 0.01),
    labels = scales::percent
  ) +
  # This makes sure that 1% on the x-axis is the same as 1% on the y-axis
  coord_fixed() +
  # This allows us to compare black v. white and Hispanic v. white side by
  # side, in panels
  facet_grid(. ~ minority_race)
  # Depending on your version of ggplot2, you may be able to use the syntax 
  # below (the newer ggplot2 syntax) -- which is clearer, in my opinion.
  # But older versions of ggplot2 will only accept the above syntax
  # facet_grid(cols = vars(minority_race))
```

This is a good start. However, it's hard to tell where the mass is. That is, 
maybe the points above the dotted line (i.e., the points where minority hit
rates are _higher_ than white hit rates), maybe those districts have all of 
Philadelphia's population, and the districts below the line only represent 
a few people. While this is unlikely, it's still good to have a marker of how 
much weight or emphasis to give each of the points on our plot.

@@

hit_rates %>% 
  left_join(
    search_counts, 
    by = c("district", "minority_race")
  ) %>% 
  ggplot(aes(
    x = white_hit_rate,
    y = minority_hit_rate
  )) +
  geom_point(aes(size = num_searches), pch = 21) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_x_continuous("White hit rate", 
    limits = c(0, max_hit_rate + 0.01),
    labels = scales::percent
  ) +
  scale_y_continuous("Minority hit rate", 
    limits = c(0, max_hit_rate + 0.01),
    labels = scales::percent
  ) +
  coord_fixed() +
  facet_grid(. ~ minority_race)
  # same facet_grid syntax issue here as before!
  # facet_grid(cols = vars(minority_race))
```

@@

(Note that we distinguish here between _sunset_ and _dusk_. Sunset is the point 
in time where the sun dips below the horizon. Dusk, also known as the end of 
civil twilight, is the time when the sun is six degrees below the horizon, and 
it is widely considered to be dark. We'll explain why this is relevant in a
moment.)

@@

One quick note: We want to make sure to filter out stops that occurred in the
approximately 30-minute period between sunset and dusk (end of civil twilight),
since it is ambiguous as to whether that period is light or dark, which would
muddle up our analysis.

@@

The veil-of-darkness test is a popular technique for assessing disparate treatment, 
but, like all statistical methods, it comes with caveats. (Hopefully that's one of
the takeaways today -- data and statistical analysis can be very powerful, but it's
good to know what the limitations are.)
Perhaps most importantly, darkness--after adjusting for time of day--is a 
function of the date. 
As such, to the extent that driver behavior changes throughout the year, and these 
changes are correlated with race, the test can suggest discrimination where there is 
none. 
One way to account for seasonal effects is to consider the brief period around daylight
savings. Driving behavior likely doesn't change much at, say, 6pm on the day before
daylight savings to the day after; but the sky is light on one day and dark on the next.
Clever checks like this can help circumvent some of the concerns about the veil of 
darkness test. However, if driving behavior is more related to lighting than time of day, the daylight savings test wouldn't help.

Another thing to note is that artificial lighting (e.g., from street lamps) can weaken the relationship between sunlight and visibility, and so the method may underestimate the extent to which stops are predicated on perceived race.

Other things, like vehicle make, year, and model often correlate with race and are
still visible at night, which could lead to the test under-estimating the extent of
racial profiling. Similarly, the test doesn't control for stop reason, which is often
correlated with both race and time of day (consider, for example, pretextual broken 
tail light stops).

Finally, the test only speaks to presence of racial profiling in the intertwilight period -- doesn't say anything about other hours.

Nevertheless, despite these shortcomings, the test provides a useful if imperfect measure of bias in stop decisions.

## Parting shots

Hopefully this gets you started on the basics of how to analyze data, and gives
you a little glimpse of some more complex modeling you can do with the data too.
While these analyses (from simple hit rates to modeling racial
profiling) can be really helpful in quantifying racial disparities, it's often
basic exploration of the data that yields the most interesting stories. So, the 
number one rule in working with data is **be curious!**

Make sure to read the [full paper](https://5harad.com/papers/100M-stops.pdf) for a 
more in-depth analysis. And we look forward to reading about all of the analysis
you all do, and all the interesting stories you uncover!
